Predicting and Preventing Employee Turnover through HR Analytics

### 2024-06-22 products/manuscript/Manuscript.qmd
Project part 3 - second update

-   What strategies can be implement to enhance employee retention and improve the overall employee experience?

@@

The dataset for this project was sourced from [Kaggle.com](https://www.kaggle.com/datasets/pavansubhasht/ibm-hr-analytics-attrition-dataset) and was created by IBM Data Scientists as a fictional representation of HR data. It consists of 1,470 rows and 35 variables that cover demographic details, job characteristics, compensation, performance metrics, and employee satisfaction data. This data serves as the foundation for analyzing factors contributing to employee turnover and developing a predictive model.

@@

In order to start the cleaning process, we need make to sure we load the data. In the code below, we use the *library(here)* function to set the file path dynamically, so our script can find and access the data files regardless of the current working directory.

@@

```{r, results='hide'}
hr_data = here::here("data","raw-data","IBM HR Analytics Data.xlsx")
rawdata = readxl::read_excel(hr_data)
```

@@

```{r, results='hide'}
d1 = rawdata %>% 
  select(-DailyRate, -HourlyRate, -MonthlyRate)

head(d1)
```

@@

```{r, results='hide'}
tenure_breaks <- seq(0, 40, by = 5)
tenure_labels <- paste0(tenure_breaks[-length(tenure_breaks)], "-", tenure_breaks[-1], " years")

d3 = d2 %>%
  mutate(TenureCategory = cut(YearsAtCompany, breaks = tenure_breaks, labels = tenure_labels, include.lowest = TRUE))
```

@@

```{r, results='hide'}
HRprocesseddata = d2

save_data_location = here::here("data","processed-data","HRprocesseddata.rds")
saveRDS(HRprocesseddata, file = save_data_location)
```

@@

I am starting my exploratory data analysis project with the objective of summarizing the dataset. I will begin by providing a concise overview of the variables included. Additionally, I will generate plots and tables to identify any outliers or patterns within the dataset. This analysis will help us understand the data available to us and guide our strategy to predict employee attrition within the organization.

@@

This dataset contains 35 variables, consisting of both character and integer types, with several character variables categorized as categorical. It provides valuable insights into employees, including demographic information such as age, gender, and marital status. Additionally, it offers details about their employment, including monthly income, department, and tenure, which are crucial for understanding their professional profiles and organizational dynamics.

@@

I examined the distribution of employee tenure and identified two distinct peaks: one at 5 years and another at 1 year. This provide quick insights into the typical duration of employees within the organization, and help us identify patterns that may influence employees to resign.

@@

The dataset consists of 1470 employees, with a gender distribution showing that males constitute the majority at 60%, while females make up 40%.

@@

## Correlation Analysis

As we explore various analysis methods in our class, my focus is on finding potential correlation between variables and employee attrition. I am interested in using the demographic details, job attributes and survey data to uncover pattern (if any) that highlight this issue .

```{r}

## Predictive Modeling

My main goal is to predict when an employee might leave the company. I am interested in using predictive modeling techniques (also part of my other summer class) to see if we can predict attrition. Once I've identified potential departures, we can use this findings to create strategies to prevent or reduce attrition.

>>>>

-   What strategies can we implement to enhance employee retention and improve the overall employee experience?

@@

The dataset for this project was sourced from [Kaggle.com](https://www.kaggle.com/datasets/pavansubhasht/ibm-hr-analytics-attrition-dataset) and was created by IBM Data Scientists as a fictional representation of HR data. It consists of 1,470 observations and 35 variables that cover demographic details, job characteristics, compensation, performance metrics, and employee satisfaction data. This data serves as the foundation for analyzing factors contributing to employee turnover and developing a predictive model.

@@

In order to start the cleaning process, we need make to sure we load the data. In the code below, we use the *library(here)* function to set the file path dynamically, so our script can find and access the data files regardless of the current working directory. You can update this code to point to a specific folder path if needed but this library will make it easier to reproduce the file with the existing folders.

@@

```{r, results='hide'}
hr_data = here::here("data","raw-data","IBM HR Analytics Data.xlsx") #file location
rawdata = readxl::read_excel(hr_data) #reading the file
```

@@

```{r, results='hide'}
d1 = rawdata %>% 
  select(-DailyRate, -HourlyRate, -MonthlyRate) # the (-) will remove these variables from the dataset

head(d1) #priting the first 6 observations
```

@@

```{r, results='hide'}
tenure_breaks <- seq(0, 40, by = 5) #defining breaks for categories
tenure_labels <- paste0(tenure_breaks[-length(tenure_breaks)], "-", tenure_breaks[-1], " years")

d3 = d2 %>%
  mutate(TenureCategory = cut(YearsAtCompany, breaks = tenure_breaks, labels = tenure_labels, include.lowest = TRUE)) #creating a new column
```

@@

```{r, results='hide'}
HRprocesseddata = d2 #creating final object

save_data_location = here::here("data","processed-data","HRprocesseddata.rds") #using the here library for the location of new file
saveRDS(HRprocesseddata, file = save_data_location)
```

@@

I am starting my exploratory data analysis project with the objective of summarizing the dataset. I will begin by providing a concise overview of the variables included. Additionally, I will generate plots and tables to identify any outliers or patterns within the dataset. This analysis will help us understand the data available to us and guide our strategy to predict employee attrition within the organization. The variables shown are derived from the dataset that we cleaned in the previous step. This includes new columns, updated values, and excludes columns that we removed.

@@

This dataset contains 33 variables, consisting of both character and integer types, with several character variables categorized as categorical. It provides valuable insights into employees, including demographic information such as age, gender, and marital status. Additionally, it offers details about their employment, including monthly income, department, and tenure, which are crucial for understanding their professional profiles and organizational dynamics.

@@

I examined the distribution of employee tenure and identified two distinct peaks: one at 5 years and another at 1 year. This provide quick insights into the typical duration of employees within the organization, and help us identify patterns that may influence employees to resign. We could concentrate on these groups to identify commonalities and understand the reasons behind their departure from the organization.

@@

The dataset consists of 1470 employees, with a gender distribution showing that males constitute the majority at 60%, while females make up 40%. This also highlights the gender disparity, and it would be interesting to explore this gap by department and examine attrition rates between the two groups.

@@

Some of these variables may explain why employees are leaving. Understanding the employee profile by examining age, gender, department, and job satisfaction can reveal the reasons behind their departure. By creating profiles based on these insights, we can develop targeted retention strategies to address the specific needs and concerns of these employees.

## Additional Analysis
